### 2024-08-15 parts/starfree.tex
enormous rewrite

\begin{lemma}
    \label{decompose-polynomial:lem}
    Let $f \colon \Sigma^* \to \Rel$ be a \kl{commutative}
    \kl{$\Rel$-polyregular function},
    where we fix the alphabet $\Sigma = \set{a_1, \dots, a_n}$.
    There exists a computable
    $\omega \in \Nat_{\geq 1}$,
    and computable 
    \kl{integer binomial polynomials} $\seqof{P_{t}}{t \in \ModuloTypes[\omega]^n}$
    such that for all $x_1, \dots, x_n \geq 0$:
    \begin{equation*}
        f\left(a_1^{x_1} \cdots a_n ^{x_n}\right) 
        = P_{\moduloType[\omega](x_1, \dots, x_n)}
        \left(
            \floor{x_1/\omega}, \dots, \floor{x_n/\omega}
        \right)
        \quad .
    \end{equation*}
    \proofref{decompose-polynomial:lem}
\end{lemma}

@@

\begin{definition}
    \label{k-combinatorial:def}
    Let $k \in \Nat$, and $f \colon \Sigma^* \to \Rel$
    be a \kl{$\Rel$-polyregular function}. The function $f$ is 
    \intro{$(k,\Nat)$-combinatorial} if there exists $\omega \in \Nat$,
    such that
    for all
    \kl{polynomial functions} $q \colon \Nat^k \to \Sigma^*$,
    there exists a polynomial $P \in \CorrectPoly$,
    satisfying:
    \begin{equation*}
        f \circ q(\omega X_1,\dots, \omega X_k)
        = 
        P
        \quad 
        \text{ over } (\Nat_{\geq 1})^k
        \quad .
    \end{equation*}
\end{definition}

@@

    Now, consider an \kl{$\omega$-type} $t \in \ModuloTypes[\omega]^k$. By
    definition, this provides us with a selection function $\ind{S} \colon \set{1,
    \dots, k} \to \set{0,1}$, and numbers $r_1, \dots, r_k$ between $0$ and
    $\omega^2-1$ such that for all $x_1, \dots, x_k \geq 1$, the tuple $(x_1
    \omega \ind{S}(1) + r_1, \dots, x_k \omega \ind{S}(k) + r_k)$ has type $t$.
    In particular, for all $x_1, \dots, x_k \geq 1$, the following
    equality holds:
    \begin{equation*}
        f\left(\prod_{i = 1}^k 
            a_i^{ x_i \omega \ind{S}(i) + r_i}
        \right)
        =
        P_t(\seqof{x_i}{i \in S})
    \end{equation*}

@@

\begin{conjecture}
    \label{npoly-zpoly:conjecture}
    Let $k \in \Nat$.
    A function $f \in \ZPoly[k]$
    belongs to $\NPoly[k]$ if and only if
    it is \kl{$(k,\Nat)$-combinatorial}.
    In particular,
    $\NPoly[k] = \ZPoly[k] \cap \NPoly$.
\end{conjecture}

Let us now prove that the above characterizations of \kl{commutative}
\kl{$\Nat$-polyregular functions} can be combined with the recent advances in
the study of \kl{$\Rel$-polyregular functions} \cite{CDTL23} to decide
aperiodicity. The key ingredient of this study is the use of a semantic
characterization of \kl{star-free $\Rel$-polyregular functions} among
\kl{$\Rel$-rational series} that generalizes the aperiodicity of languages to
functions by the means of polynomial behaviors (see
\cref{aperidic-ultimately-polynomial:ex}). Note that up until now, we only had
to consider polynomials with coefficients in $\Rel$. However, there are some
polynomials with coefficients in $\Rat$ that can be \kl{represented} by a
\kl{$\Nat$-polyregular function}. Recall that the polynomial $P(X) \defined
X(X-1)/2$ can be computed by a \kl{$\Nat$-polyregular function}. Historically,
the definition used polynomials with coefficients in $\Rat$, which is why we
use such polynomials here, even though we know that all of those polynomials
will actually be \kl{integer binomial polynomials} thanks to
\cref{integer-binomial-polynomials:cor}.

@@

    Conversely, assume that $L$ is aperiodic, let $n \in \Nat$ be a number, and
    let $\alpha_0$, $w_1$, $\alpha_1$, $\ldots$, $\alpha_{n-1}$, $w_n$, $\alpha_n$
    be words in $\Sigma^*$. Because $L$ is \kl(language){aperodic}, the language
    $L$ is recognized by an \kl(monoid){aperiodic} finite monoid $M$, via a
    morphism $\mu \colon \Sigma^* \to M$ and an accepting part $F \subseteq M$.
    Because $M$ is a \kl(monoid){aperiodic}, there exists $N_0$
    such that for all $m \in M$, $m^{N_0+1} = m^{N_0}$. As a consequence,
    for all $1 \leq i \leq n$,
    whenever $x_i \geq N_0$, we have $\mu(w_i)^{x_i} = \mu(w_i)^{N_0}$.
    Therefore, for all $x_1, \dots, x_n \geq N_0$, we have:
    \begin{align*}
        \ind{L}\left(\alpha_0 \prod_{i= 1}^n w_i^{x_i} \alpha_i \right) 
        &= 
        \ind{P}\left(\mu(\alpha_0) \cdot \prod_{i= 1}^n \mu(w_i)^{x_i} \cdot \mu(\alpha_i)\right)
        & \text{ because $M$ recognizes $L$}
        \\
        &= 
        \ind{P}\left(\mu(\alpha_0) \cdot \prod_{i= 1}^n \mu(w_i)^{N_0} \cdot \mu(\alpha_i)\right)
        & \text{ by aperiodicity }
        \\
        &\defined P(x_1, \dots, x_n) & \text{ constant polynomial }
    \end{align*}
    Which proves that $\ind{L}$ is \kl{ultimately polynomial}.
\end{proof}

>>>>

\begin{lemma}
    \label{decompose-polynomial:lem}
    \proofref{decompose-polynomial:lem}
    Let $f \colon \Sigma^* \to \Rel$ be a \kl{commutative}
    \kl{$\Rel$-polyregular function},
    where we fix the alphabet $\Sigma = \set{a_1, \dots, a_n}$.
    There exists a computable
    $\omega \in \Nat_{\geq 1}$,
    and computable 
    \kl{integer binomial polynomials} $\seqof{P_{t}}{t \in \ModuloTypes[\omega]^n}$
    such that for all $x_1, \dots, x_n \geq 0$:
    \begin{equation*}
        f\left(a_1^{x_1} \cdots a_n ^{x_n}\right) 
        = P_{\moduloType[\omega](x_1, \dots, x_n)}
        \left(
            \floor{x_1/\omega}, \dots, \floor{x_n/\omega}
        \right)
        \quad .
    \end{equation*}
\end{lemma}

@@

\begin{definition}
    \label{k-combinatorial:def}
    Let $k \in \Nat$, and $f \colon \Sigma^* \to \Rel$
    be a \kl{$\Rel$-polyregular function}. The function $f$ is 
    \intro{$(k,\Nat)$-combinatorial} if there exists $\omega \in \Nat$,
    such that
    for all
    \kl{polynomial functions} $q \colon \Nat^k \to \Sigma^*$,
    there exists a polynomial $P \in \CorrectPoly$,
    satisfying:
    \begin{equation*}
        f \circ q(\omega X_1,\dots, \omega X_k)
        = 
        P
        \quad 
        \text{ over } (\Nat_{\geq 1})^k
        \quad .
    \end{equation*}
\end{definition}
\begin{conjecture}
    \label{npoly-zpoly:conjecture}
    Let $k \in \Nat$.
    A function $f \in \ZPoly[k]$
    belongs to $\NPoly[k]$ if and only if
    it is \kl{$(k,\Nat)$-combinatorial}.
    In particular,
    $\NPoly[k] = \ZPoly[k] \cap \NPoly$.
\end{conjecture}

@@

    Now, consider $t \in \ModuloTypes[\omega]^k$. By
    definition, this provides us with a set $S \subseteq \set{1, \dots, k}$,
    and numbers $r_1, \dots, r_k$ between $0$ and
    $\omega^2-1$ such that for all $x_1, \dots, x_k \geq 1$, the tuple $(x_1
    \omega \ind{S}(1) + r_1, \dots, x_k \omega \ind{S}(k) + r_k)$ has type $t$.
    In particular, for all $x_1, \dots, x_k \geq 1$, the following
    equality holds:
    \begin{equation*}
        f\left(\prod_{i = 1}^k 
            a_i^{ x_i \omega \ind{S}(i) + r_i}
        \right)
        =
        P_t(\seqof{x_i}{i \in S})
    \end{equation*}

@@

Let us now prove that the above characterizations of \kl{commutative}
\kl{$\Nat$-polyregular functions} can be combined with the recent advances in
the study of \kl{$\Rel$-polyregular functions} \cite{CDTL23} to decide
aperiodicity. The key ingredient of this study is the use of a semantic
characterization of \kl{star-free $\Rel$-polyregular functions} among
\kl{$\Rel$-rational series} that generalizes the aperiodicity of languages to
functions by the means of polynomial behaviors (see
\cref{aperidic-ultimately-polynomial:ex}). Historically, the definition used
polynomials with coefficients in $\Rat$, which is why we use such polynomials
here, even though we know that all of those polynomials will actually be
\kl{integer binomial polynomials} thanks to
\cref{integer-binomial-polynomials:cor}.

@@

    Conversely, assume that $L$ is \kl(language){aperiodic}, let $n \in \Nat$ be a number, and
    let $\alpha_0$, $w_1$, $\alpha_1$, $\ldots$, $\alpha_{n-1}$, $w_n$, $\alpha_n$
    be words in $\Sigma^*$. Because $L$ is \kl(language){aperiodic}, the language
    $L$ is recognized by an \kl(monoid){aperiodic} finite monoid $M$, via a
    morphism $\mu \colon \Sigma^* \to M$ and an accepting part $F \subseteq M$.
    Because $M$ is a \kl(monoid){aperiodic}, there exists $N_0$
    such that for all $m \in M$, $m^{N_0+1} = m^{N_0}$. As a consequence,
    for all $1 \leq i \leq n$,
    whenever $x_i \geq N_0$, we have $\mu(w_i)^{x_i} = \mu(w_i)^{N_0}$.
    Therefore, for all $x_1, \dots, x_n \geq N_0$, we have:
    \begin{align*}
        \ind{L}\left(\alpha_0 \prod_{i= 1}^n w_i^{x_i} \alpha_i \right) 
        &= 
        \ind{P}\left(\mu(\alpha_0) \cdot \prod_{i= 1}^n \mu(w_i)^{x_i} \cdot \mu(\alpha_i)\right)
        & \text{ because $M$ recognizes $L$}
        \\
        &= 
        \ind{P}\left(\mu(\alpha_0) \cdot \prod_{i= 1}^n \mu(w_i)^{N_0} \cdot \mu(\alpha_i)\right)
        & \text{ by aperiodicity }
        \\
        &\defined P(x_1, \dots, x_n) & \text{ constant polynomial }
    \end{align*}
    Which proves that $\ind{L}$ is \kl{ultimately polynomial}.
\end{proof}
